Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-06-11 newpaper/paper.tex
stuff

\textbf{Two-level type theory} (2LTT) \cite{twolevel,staged2ltt} makes it
possible to use expressive dependent type theories for metaprogramming
for a wide range of object languages. In this paper, we use 2LTT to sweeten the
deal of staged compilation, aiming for a combination of strong guarantees, good
ergonomics, high level of abstraction and easy-to-optimize code output.

We develop a particular two-level type theory for this purpose, which we
call \textbf{CFTT}, short for ``closure-free type theory''. This consists
of:
\begin{itemize}
\item A simply-typed object theory with first-order functions, general recursion and
      finitary algebraic data types. This language is easy to optimize and compile
      in the downstream pipeline, but it lacks many convenience features.
\item A dependent type theory for the compile-time language. This
      allows us to recover many features by metaprogramming.
\end{itemize}
Since the object language is first-order, we guarantee that all programs in CFTT
can be ultimately compiled without any dynamic closures, using only calls and
jumps to statically known code. Why emphasize closures?  They are the
foundation to almost all abstraction tools in functional programming:
\begin{itemize}
\item Higher-order functions in essentially all functional languages are implemented with closures.
\item Type classes in Haskell use dictionary-passing, which relies on closures for function methods \cite{DBLP:conf/popl/WadlerB89}.
\item Functors and first-class modules in OCaml \cite{leroy2023ocaml} and other ML-s rely on closures.
\end{itemize}
Hence, doing functional programming without closures is a clear demonstration
that we can get rid of abstraction overheads.

@@

Formally, $\MTy$ is additionally indexed by universes levels (orthogonally to
staging), and we have $\MTy_i : \MTy_{i+1}$. However, universe levels add noise
and they are not too relevant to the current paper, so we will omit them.

@@

Object-level definitions use $:=$ as notation, instead of the $=$ that is used
for meta-level ones. We also have non-recursive $\mbf{let}$, which can be used
to define computations and values alike, and can be used to shadow binders:
\begin{alignat*}{3}
  &\vf : \Nat \to \Nat\\
  &\vf\,\vx := \letdef \vx := \vx + 10; \letdef \vx := \vx + 20; \vx * 10
\end{alignat*}
We also allow $\mbf{newtype}$ definitions, both in $\VTy$ and $\CTy$. These are
assumed to be erased at runtime. In the Haskell implementation they are
important for guiding type class resolution, and we think that the explicit
wrapping makes many definitions more comprehensible in CFTT as well.

@@

\subsubsection{Comparison to defunctionalization}
Defunctionalization is a program translation which represents higher-order
functions using only first-order functions and inductive data
\cite{DBLP:journals/lisp/Reynolds98a,DBLP:conf/ppdp/DanvyN01}. The idea is to
represent each closure as a constructor of an inductive type, and implement
closure application as a top-level first-order function which switches on all
possible closure constructors. Hence, like our unstaging algorithm,
defunctionalization produces first-order code from higher-order code. We summarize
the differences.
\begin{itemize}
\item Defunctionalization mainly supports \emph{compiler optimizations}, by making
  closures transparent to analysis. It is not an optimization by itself; it does
  not change the amount of dynamic control flow or dynamic allocations in a
  program. A dynamic closure call becomes a dynamic case-switch and a
  heap-allocated ``functional'' closure becomes a heap-allocated inductive
  constructor.
\item Staging mainly supports optimization by \emph{programmers} by providing
  control over code generation. Unstaging does not translate higher-order
  functions to anything in particular. Instead, it is the execution of
  higher-order metaprograms that produce first-order programs.
\end{itemize}

@@

We give an overview of the unstaging algorithm together with its correctness.
Our main task is to extend \cite{staged2ltt} with identity types and inductive
types in the metalanguage, since we use them in this paper. For inductive types,
it is the most sensible to add W-types, since all inductive families can be
faithfully derived from them in our setting \cite{whynotw}.

>>>>

We use \textbf{two-level type theory} (2LTT) \cite{twolevel,staged2ltt} to
sweeten the deal of staged compilation, aiming for a combination of strong
guarantees, good ergonomics, high level of abstraction and easy-to-optimize code
output. We develop a particular two-level type theory for this purpose, which we
call \textbf{CFTT}, short for ``closure-free type theory''. This consists of:
\begin{itemize}
\item A simply-typed object theory with first-order functions, general recursion and
      finitary algebraic data types. This language is easy to optimize and compile
      in the downstream pipeline, but it lacks many convenience features.
\item A dependent type theory for the compile-time language. This
      allows us to recover many features by metaprogramming.
\end{itemize}
Since the object language is first-order, we guarantee that all programs in CFTT
can be ultimately compiled without any dynamic closures, using only calls and
jumps to statically known code. Why emphasize closures?  They are the
foundation to almost all abstraction tools in functional programming:
\begin{itemize}
\item Higher-order functions in essentially all functional languages are implemented with closures.
\item Type classes in Haskell use dictionary-passing, which relies on closures for function methods \cite{DBLP:conf/popl/WadlerB89}.
\item Functors and first-class modules in OCaml \cite{leroy2023ocaml} and other ML-s rely on closures.
\end{itemize}
Hence, doing functional programming without closures is a clear demonstration
that we can get rid of abstraction overheads.

@@

Formally, $\MTy$ is additionally indexed by universe levels (orthogonally to
staging), and we have $\MTy_i : \MTy_{i+1}$. However, universe levels add noise
and they are not too relevant to the current paper, so we will omit them.

@@

Object-level definitions use $:=$ as notation, instead of the $=$ that is used
for meta-level ones. We also have non-recursive $\mbf{let}$, which can be used
to define computations and values alike, and can be used to shadow binders:
\begin{alignat*}{3}
  &\vf : \Nat \to \Nat\\
  &\vf\,\vx := \letdef \vx := \vx + 10; \letdef \vx := \vx + 20; \vx * 10
\end{alignat*}
We also allow $\mbf{newtype}$ definitions, both in $\VTy$ and $\CTy$. These are
assumed to be erased at runtime. In the Haskell and Agda implementations they
are important for guiding instance resolution, and we think that the explicit
wrapping makes many definitions more comprehensible in CFTT as well.

@@

\subsubsection{Comparison to defunctionalization}
Defunctionalization is a program translation which represents higher-order
functions using only first-order functions and inductive data
\cite{DBLP:journals/lisp/Reynolds98a,DBLP:conf/ppdp/DanvyN01}. The idea is to
represent each closure as a constructor of an inductive type, and implement
closure application as a top-level first-order function which switches on all
possible closure constructors. Hence, like our unstaging algorithm,
defunctionalization produces first-order code from higher-order code. We summarize
the differences.
\begin{itemize}
\item Defunctionalization mainly supports \emph{compiler optimizations}, by making
  closures transparent to analysis. It is not an optimization by itself; it does
  not change the amount of dynamic control flow or dynamic allocations in a
  program. A dynamic closure call becomes a dynamic case-switch and a
  heap-allocated ``functional'' closure becomes a heap-allocated inductive
  constructor.
\item Staging mainly supports optimization by \emph{programmers} by providing
  control over code generation. Unstaging does not translate higher-order
  functions to anything in particular. Instead, unstaging is the execution of
  higher-order metaprograms which produce first-order programs.
\end{itemize}

@@

To obtain an unstaging algorithm, together with its correctness, the main task
is to extend \cite{staged2ltt} with identity types and inductive families in the
metalanguage, since we use those types in this paper. For inductive types, it is the
most sensible to add W-types, since all inductive families can be faithfully
derived from them in our setting \cite{whynotw}.
